What is essential? -- A pilot survey on views about the requirements metamodel of reqT

### 2015-10-01 paper/paper-refsq2016-submission.tex
add analysis and update paper

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
%%%%TODO: REMOVE THE TERM metacompleteness
%% MY COOL PACKAGES
\usepackage{times}
\usepackage[normalem]{ulem}
\usepackage{cite}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{scalefnt}
\usepackage{setspace}
%
%\usepackage{listings}
%\lstset{basicstyle=\ttfamily\small, 
%%  numbers=left, numberstyle=\tiny, stepnumber=1, numbersep=5pt,
%  language=java, mathescape=true, escapechar=\!,
%  emph={object}, emphstyle=\textbf,
%  showstringspaces=false
%}
%%%%

\usepackage{url}
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

\begin{document}

@@

%%%% ABSTRACT
\begin{abstract}
[{\bf Context and motivation}] This research preview presents ongoing work on the metamodel of a free software requirements modeling tool called reqT that is developed in an educational context. [{\bf Question/problem}] The work aims to make an initial validation of a survey instrument that elicits views on the metamodel of the reqT tool, which seek to engage computer science students in Requirements Engineering (RE) through an open source requirements engineering DSL embedded in the Scala programming language. The research question underpinning the presented work is: Which RE concepts are essential to include in the metamodel for a requirements engineering tool in an educational context?  [{\bf Principal ideas}] A survey instrument is developed with a list of 92 concepts (49 entities, 15 relations and 28 attributes) and a set of questions for each concept that elicit the respondents' views on the usage and interpretation of each concept.  [{\bf Contribution}] The survey is initially validated in a pilot study involving 14 Swedish RE scholars as subjects. The survey results indicate that the survey is feasible if the respondents is willing to invest around 30 minutes of their time. The analysis of the responses suggest that many of the concepts in the metamodel are used frequently by the respondents and there is a large degree of agreement among the respondents about the meaning of the concepts. The results are encouraging for future work on empirical validation of the relevance of the reqT metamodel.

@@

%%%%% INTRO
\section{Introduction}
There are many challenges in teaching Requirements Engineering (RE) \cite{Memon2010, Regev2011}, including  conveying requirements modeling skills \cite{Callele2006}. Given a wide-spread attention on agile methods with less emphasis on extra-code artifacts \cite{Ramesh2010}, it may be particularly challenging to motivate coding-focused engineering students (and software practitioners) to spend serious effort on requirements modeling. One way to inspire software engineers to learn more about and do more RE may be to offer an interesting software tool. There are nowadays numerous commercial RE tools available, but many are expensive, complex and not sufficiently open  \cite{Carillo2011}.

This paper presents on-going work on a tool named reqT that aims to provide a small but scalable, semi-formal and free software package for an educational setting that (hopefully) can inspire code lovers to learn more about requirements modeling. A long-term goal of reqT is to offer an open platform for RE research prototypes, e.g. for feature modeling and release planning research. The tool development started in 2011 at Lund University, where reqT is used in RE teaching at MSc level in the Computer Science \& Engineering program. In 2012 reqT was rebuilt from scratch based on student feedback. The tool can be downloaded from: \url{http://reqT.org}

The paper is organized as follows. Section \ref{section:objectives} states the objectives and motivates the design strategy of reqT. Section \ref{section:metamodel} presents the metamodel of reqT and some example reqT models. Section \ref{section:discussion} discusses limitations and some initial experiences from using reqT in teaching and concludes the paper with a sketch of future research directions.

\section{Goals, Design Strategy and Rationale}\label{section:objectives}
The main objective behind reqT is to establish a set of essential RE concepts and capture them in an expressive, extensible and executable language appealing to computer science students (and eventually practitioners). This general objective is accompanied by the following main goals and associated design strategies:
\begin{enumerate}
\item {\bf Semi-formal.} {\it Goal:} Provide a semi-formal representation of typical requirements modeling constructs that can illustrate a flexible combination of expressive natural language-style requirements with type-safe formalisms allowing static checks. {\it Design:} Use graph structures based on typed nodes representing typical requirement entities and attributes, and typed edges representing typical requirements relations, and implement the graph as an associative array (map). {\it Why?} Graphs are well-known to many CS students. Maps are efficient from an implementation perspective and may be less complex to master compared to e.g. SQL databases.
\item {\bf Open.} {\it Goal:} Provide a platform-independent requirements tool that is free of charge. {\it Design:} Use Java Virtual Machine technology and release the code under an open source license. Use tab-separated, tabular text-files for import and export. Use HTML for document generation. {\it Why?} There are many free libraries available that runs on a JVM. Tab-sep and HTML support interoperability. %A free, platform-independent tool is appreciated by many students.
\item {\bf Scalable.} {\it Goal:}  Provide an extensible requirements modeling language that can scale from small, concise models to large families of models with thousands of requirements entities and relations. {\it Design:} Implement reqT as an internal DSL (Domain-Specific Language) in the Scala programming language \cite{Odersky2004}. Use Map and Set from Scala collections to represent requirements graphs. {\it Why?} Scala is a modern, statically typed language with an efficient collections library. Scala offers scripting abilities that provide general extensibility without re-compilation. Integrated development environments \cite{eclipse}, as well as interactive scripting tools are available \cite{kojo}. %When requirements are expressed as code, general software engineering tools can be applied when scaling to large projects, e.g. configuration management tools and cloud-based code repositories to facilitate requirements evolution and collaboration over Internet. 
\end{enumerate}

These goals, design strategies and rationale are directing the on-going work, and it remains to be investigated to what extent the main objective and goals can be met. A critical issue is how to interpret what are "essential" RE concepts and "typical" modeling constructs. The reqT tool is used in a course based on a specific text book \cite{Lauesen2002} and a specific student project concept \cite{ets170}, and the concepts of the reqT requirements metamodel (see Fig. \ref{fig:types}) reflect that context. However, the reqT architecture is prepared for extensions of new concepts in the metamodel to cater for different educational contexts. 
%%%%%%% THE MODEL
\section{Modeling requirements with reqT}\label{section:metamodel}

A reqT model includes sequences of graph parts {\tt<Entity><Edge><NodeSet>}  separated by comma and wrapped inside a {\tt Model( )} construct. A small reqT Model with three {\tt Feature} entities and one {\tt Stakeholder} entity is shown below:
\begin{spacing}{0.8}
\begingroup
    \fontsize{8pt}{12pt}\selectfont
\begin{verbatim}
Model(
  Feature("f1") has (Spec("A good spec."), Status(SPECIFIED)),
  Feature("f1") requires (Feature("f2"), Feature("f3")),
  Stakeholder("s1") assigns(Prio(1)) to Feature("f2")
)
\end{verbatim}
\endgroup
\end{spacing}

%%%%%%%%%%%%%% DISCUSSION

@@

 The results of the on-going work with reqT remains to be further investigated and a validation of reqT as a RE learning tool and research experimentation platform is subject to future work. This section discusses some preliminary experiences, limitations, relation to state-of-the-art and future research directions.

{\bf Preliminary proof-of-concept.}
The first version of reqT was tried on a voluntary basis by 12 students working in groups of 6 students each during fall 2011. Statements from course evaluations indicate that the students found reqT useful in their learning. One group used a configuration management tool for reqT models to manage  their parallel work, while one group used a cloud service and tab-sep export/import to collaborate over the Internet. The group with the largest requirements model produced 64 features, 18 tasks, 12 functions, 30 data requirements and 33 quality requirements, in total 157 requirements entities.

Several students appreciated that reqT can mix informal text with a graph-oriented formalism, but some requested more elaborated functionality for document generation, as well as linking to external images. Some students also requested more modeling examples that show how the text book techniques could be transferred to reqT models.

Based on student feedback, reqT was rebuilt from scratch during 2012 with a new architecture and a new version of the meta model (see Fig. \ref{fig:types}), as well as a revised Scala-internal DSL. The template-controlled HTML generation was implemented based on student suggestions. The teaching material was complemented with more example models directly related to the textbook. The second version of reqT is currently tested by  students in a new course instance and a post-course evaluation of reqT is planned in spring 2013.

Our preliminary experiences from applying reqT in teaching suggest that reqT, if used in a suitable teaching context, may encourage students with a code-focused mind set to learn and practice RE in the following ways: (1) A free and platform-independent software tool that is implemented using a modern programming language with interactive scripting facilities can attracts the interest of code-focused students. (2) Requirements can be processed, queried, transformed or exported using Scala scripts, and the open-ended nature of reqT that allows students to code their own scripts to both manage requirements models and to adapt reqT to fit their RE needs in the course project was appreciated by several coding-literate students. (3) By turning requirements models into executable code, students can use programming tools such as a console command line interpreter (the Scala REPL) as well as a source code version control system (e.g. git-scm.com) to branch and merge their collaborative work on requirements in ways they are used to from their previous collaborative software implementation courses, including issue tracking systems and code review support.

{\bf Relation to state-of-the-art.} To the best of our knowledge there is no other RE tool that allows semi-formal requirement models to become executable programs through an internal Scala DSL, and thus letting coding, testing and requirements engineering share the same media. In the RE tool survey by Carrillo de Gea et al.  \cite{Carillo2011} it is pointed out that "many expensive tools aren't sufficiently open". The reqT technology aims to be completely free and open to facilitate academic usage, collaborative evolution and incorporation of new RE concepts in different teaching and research contexts. Many of the existing tools have proprietary representations \cite{Carillo2011}, while users of reqT can extend the reqT metamodel with new entities and attributes simply by adding case classes with a few lines of code. However, reqT cannot compete with versatile commercial RE tools  \cite{Carillo2011}  in terms of e.g. features completeness and graphical user interface.

{\bf Limitations. } In its current version, reqT has a number of limitations: (1) As the user interface is text based and depends on the command line interface of the Scala REPL or a script editor environment \cite{kojo, eclipse}, students that only are prepared to use graphical user interfaces may be discouraged. Some of our students preferred to work in a GUI spreadsheet application using tab-separated exports from reqT that was generated by other team members assigned by the student group to be reqT experts. (2) It requires some knowledge of Scala to tailor reqT exports and there is a need for a more comprehensive API for adaptable document generation. (3) The embedded DSL requires some learning efforts and it remains to be investigated if the effort is justified by the knowledge gained. (4) To support scalability to large families of reqT models there is a need for modularization concepts and overview visualizations. (5) The explicit typing of entities with keywords such Feature and Stakeholder can be perceived as verbose compared to more concise but potentially cryptic abbreviations (e.g. Fe, Sh). This may be addressed by DSL-specific editor support, such as code-completion, code folding and code templates.

{\bf Future work.} Further directions of research include (1) incorporation of constraints on models for support of prioritization and release planning \cite{Regnell2011}, (2) more elaborate semantic checks to better guide requirements modelers, and (3) graphical visualization of requirements graph models. (4) Natural Language Processing technology including e.g. ambiguity risk detection may be interesting to combine with reqT. (5) It is also important to further investigate the pedagogic advantages and limitations of the approach.

A major objective of this research preview paper is to expose the latest version of reqT to the community of RE scholars and to invite discussions and contributions.

@@

\clearpage
\hspace{-0.5cm}
\begin{minipage}{1\linewidth}
\begin{minipage}{0.36\textwidth} 
\begin{table}[H]
%\centering
\vspace{-0.7cm}
\fontsize{6}{6.5}\selectfont
\caption{Definitions of concepts.}%: Entities, Attributes and Relations.}
\label{table-entities}
\begin{tabular}{|l p{5.0cm}|}
\hline
\textit{Entity} & \textit{Definition}  \\ \hline
Actor&A human or machine that communicates with a system.\\
App&A computer program, or group of programs designed for end users, normally with a graphical user interface. Short for application.\\
Barrier&Something that makes it difficult to achieve a goal or a higher quality level.\\
Breakpoint&A point of change. An important aspect of a (non-linear) relation between quality and benefit.\\
Class&An extensible template for creating objects. A set of objects with certain attributes in common. A category.\\
Component&A composable part of a system. A reusable, interchangeable system unit or functionality.\\
Configuration&A specific combination of variants.\\
Data&Information stored in a system.\\
Design&A specific realization or high-level implementation description (of a system part).\\
Domain&The application area of a product with its surrounding entities.\\
Epic&A large user story or a collection of stories.\\
Event&Something that can happen in the domain and/or in the system.\\
Feature&A releasable characteristic of a product. A (high-level, coherent) bundle of requirements.\\
Function&A description of how input data is mapped to output data. A capability of a system to do something specific.\\
Goal&An intention of a stakeholder or desired system property.\\
Idea&A concept or thought (potentially interesting).\\
Interface&A defined way to interact with a system.\\
Issue&Something needed to be fixed.\\
Item& An article in a collection, enumeration, or series.\\
Label&A descriptive name used to identify something.\\
Member&An entity that is part of another entity, eg. a field in a in a class.\\
Meta&A prefix used on a concept to mean beyond or about its own concept, e.g. metadata is data about data.\\
MockUp&A prototype with limited functionality used to demonstrate a design idea.\\
Module&A collection of coherent functions and interfaces.\\
Product&Something offered to a market.\\
Quality&A distinguishing characteristic or degree of goodness.\\
Relationship&A specific way that entities are connected.\\
Release&A specific version of a system offered at a specific time to end users.\\
Req&Something needed or wanted. An abstract term denoting any type of information relevant to the (specification of) intentions behind system development. Short for requirement.\\
Resource&A capability of, or support for development.\\
Risk&Something negative that may happen.\\
Scenario&A (vivid) description of a (possible future) system usage.\\
Screen&A design of (a part of) a user interface.\\
Section&A part of a (requirements) document.\\
Service&Actions performed by systems and/or humans to provide results to stakeholders.\\
Stakeholder&Someone with a stake in the system development or usage.\\
State&A mode or condition of something in the domain and/or in the system. A configuration of data.\\
Story&A short description of what a user does or needs. Short for user story.\\
System&A set of interacting software and/or hardware components.\\
Target&A desired quality level or goal .\\
Task&A piece of work (that users do, maybe supported by a system).\\
Term&A word or group of words having a particular meaning.\\
Test&A procedure to check if requirements are met.\\
Ticket&(Development) work awaiting to be completed.\\
UseCase&A list of steps defining interactions between actors and a system to achieve a goal.\\
User&A human interacting with a system.\\
Variant&An object or system property that can be chosen from a set of options.\\
VariationPoint&An opportunity of choice among variants.\\
WorkPackage&A collection of (development) work tasks.\\
\hline
 \end{tabular}
\end{table}
\end{minipage}
 \hspace{4em plus 1fill} 
\begin{minipage}{0.47\textwidth} 
 \begin{table}[H]
%\centering
\fontsize{6}{6.5}\selectfont
%\caption{Definitions of attributes and relations.}
\vspace{-0.2cm}
\label{table-entities}
\begin{tabular}{|l p{4.3cm}|}
 \hline
\textit{Attribute} & \textit{Definition}  \\ \hline
Benefit&A characterisation of a good or helpful result or effect (e.g. of a feature).\\
Capacity&The largest amount that can be held or contained (e.g. by a resource).\\
Code&A collection of (textual) computer instructions in some programming language, e.g. Scala. Short for source code.\\
Comment&A note that explains or discusses some entity.\\
Constraints&A collection of propositions that restrict the possible values of a set of variables.\\
Cost&The expenditure of something, such as time or effort, necessary for the implementation of an entity.\\
Damage&A characterisation of the negative consequences if some entity (e.g. a risk) occurs.\\
Deprecated&A description of why an entity should be avoided, often because it is superseded by another entity, as indicated by a 'deprecates' relation.\\
Example&A note that illustrates some entity by a  typical instance.\\
Expectation&The required output of a test in order to be counted as passed.\\
FileName&The name of a storage of serialized, persistent data.\\
Frequency&The rate of occurrence of some entity. \\
Gist&A short and simple description of an entity, e.g. a function or a test.\\
Image&(The name of) a picture of an entity.\\
Input&Data consumed by an entity, \\
Max&The maximum estimated or assigned (relative) value.\\
Min&The minimum estimated or assigned (relative) value.\\
Order&The ordinal number of an entity (1st, 2nd, ...).\\
Output&Data produced by an entity, e.g. a function or a test.\\
Prio&The level of importance of an entity. Short for priority.\\
Probability&The likelihood that something (e.g. a risk) occurs.\\
Profit&The gain or return of some entity, e.g. in monetary terms.\\
Spec&A (detailed) definition of an entity. Short for specification\\
Status&A level of refinement of an entity (e.g. a feature) in the development process. \\
Text&A sequence of words (in natural language).\\
Title&A general or descriptive heading.\\
Value&An amount. An estimate of worth.\\
Why&A description of intention. Rationale.\\

\hline
\textit{Relation} & \textit{Definition}  \\ \hline
binds&Ties a value to an option. A configuration binds a variation point.\\
deprecates&Makes outdated. An entity deprecates (supersedes) another entity.\\
excludes&Prevents a combination. An entity excludes another entity.\\
has&Expresses containment, substructure. An entity contains another entity.\\
helps&Positive influence. A goal helps to fulfil another goal.\\
hurts&Negative influence. A goal hinders another goal.\\
impacts&Some influence. A new feature impacts an existing component.\\
implements&Realisation of. A module implements a feature.\\
interactsWith&Communication. A user interacts with an interface.\\
is&Sub-typing, specialization, part of another, more general entity.\\
precedes&Temporal ordering. A feature precedes (is implemented before) another feature.\\
relatesTo&General relation. An entity is related to another entity.\\
requires&Requested combination. An entity is required (or wished) by another entity.\\
superOf&Super-typing, generalization, includes another, more specific entity.\\
verifies&Gives evidence of correctness. A test verifies the implementation of a feature.\\
 \hline
\end{tabular}
\end{table}
\end{minipage}
\end{minipage}
\subsubsection*{Acknowledgments.} This work is partly funded by  VINNOVA within the EASE project.

>>>>

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
%%%%TODO: REMOVE THE TERM metacompleteness
%% MY COOL PACKAGES
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
%\usepackage{times}
\usepackage{tgtermes}
%\usepackage{bera}
\usepackage[scaled=0.93]{beramono}
\usepackage[normalem]{ulem}
\usepackage{cite}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{scalefnt}
\usepackage{setspace}
%
%\usepackage{listings}
%\lstset{basicstyle=\ttfamily\small, 
%%  numbers=left, numberstyle=\tiny, stepnumber=1, numbersep=5pt,
%  language=java, mathescape=true, escapechar=\!,
%  emph={object}, emphstyle=\textbf,
%  showstringspaces=false
%}
%%%%

\usepackage{hyperref}
\usepackage{xcolor}
\definecolor{entityColor}{RGB}{0,100,200}
\definecolor{attributeColor}{RGB}{0,100,50}
\definecolor{relationColor}{RGB}{160,0,30}
\usepackage{listings}
\lstdefinestyle{reqT}{
  %belowcaptionskip=1\baselineskip,
  breaklines=true,
  %showstringspaces=false,
  showspaces=false,
  %breakatwhitespace=true,
  basicstyle=\ttfamily\scriptsize,
  emph={Ent,Meta,Item,Label,Section,Term,Actor,App,Component,Domain,Module,Product,Release,Resource,Risk,Service,Stakeholder,System,User,Class,Data,Input,Member,Output,Relationship,Design,Screen,MockUp,Function,Interface,State,Event,Epic,Feature,Goal,Idea,Issue,Req,Ticket,WorkPackage,Breakpoint,Barrier,Quality,Target,Scenario,Task,Test,Story,UseCase,VariationPoint,Variant},
  emphstyle=\bfseries\color{entityColor},
  emph={[2]has,is,superOf,binds,deprecates,excludes,helps,hurts,impacts,implements,interactsWith,precedes,requires,relatesTo,verifies},
  emphstyle={[2]\color{relationColor}},
  emph={[3]Attr,Code,Constraints,Comment,Deprecated,Example,Expectation,FileName,Gist,Image,Spec,Text,Title,Why,Benefit,Capacity,Cost,Damage,Frequency,Min,Max,Order,Prio,Probability,Profit,Value,Status},
  emphstyle={[3]\itshape \color{attributeColor}},  
}
\lstset{style=reqT}

\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}
\begin{document}

@@

%%%% ABSTRACT
\begin{abstract}
[{\bf Context \& motivation}] This research preview presents ongoing work on the metamodel of a free software requirements modeling tool called reqT that is developed in an educational context. The work aims to make an initial validation of a survey instrument that elicits views on the metamodel of the reqT tool, which seek to engage computer science students in Requirements Engineering (RE) through an open source requirements engineering DSL embedded in the Scala programming language. [{\bf Question}] The research question is: Which RE concepts are essential to include in the metamodel for a requirements engineering tool in an educational context?  [{\bf Principal ideas}] A survey instrument is developed with a list of 92 concepts (49 entities, 15 relations and 28 attributes) and a set of questions for each concept that elicit the respondents' views on the usage and interpretation of each concept.  [{\bf Contribution}] The survey is initially validated in a pilot study involving 14 Swedish RE scholars as subjects. The survey results indicate that the survey is feasible if the respondents is willing to invest around 30 minutes of their time. The analysis of the responses suggest that many of the concepts in the metamodel are used frequently by the respondents and there is a large degree of agreement among the respondents about the meaning of the concepts. Some terms can be viewed as ''essential RE concepts'' in that a many use them and agree on their meaning. The results are encouraging for future work on empirical validation of the relevance of the reqT metamodel.

@@

%%%%% INTRO
\section{Introduction}
There are many challenges in teaching Requirements Engineering (RE) \cite{Memon2010, Regev2011}, including  conveying requirements modelling skills that can be used effectively in an unstructured, non-ideal, real-world situation \cite{Callele2006}. When teaching RE modelling we may ask ourselves: What are the \textit{essential} RE concepts that we should include in our taught metamodel for requirements? This paper investigates this questions in conjunction with the on-going work of developing a metamodel for reqT \cite{reqT}, an open source requirements engineering tool \cite{Regnell2013} used in RE education \cite{ets170}.
A survey instrument is presented aiming to elicit the frequency of RE term usage and the degree of interpretation agreement. The responses from 15 Swedish RE scholars are analysed and discussed and conclusions suggest that a large subset of the concepts of the current reqT metamodel can be argued to be ''essential'' in that a majority of the subjects use them while agreeing with the concept definitions. The presented work is an initial validation and further work involving more subjects is needed to provide conclusions with more certainty.

\section{Background and Related Work}

There are nowadays numerous commercial RE tools available, but many are expensive, complex and not sufficiently open  \cite{Carillo2011}. A major aim of the reqT open source project is to provide a small but scalable, semi-formal and free software package for an educational setting \cite{Regnell2013} that can inspire code-loving computer science students to learn more about requirements modeling. The tool development started in 2011 at Lund University, where reqT is used in RE teaching at MSc level in the Computer Science \& Engineering program \cite{ets170}.

A critical issue is how to find the "essential" RE concepts that allows for sufficient expressiveness, while not overloading the metamodel with esoteric concepts just for the sake of completeness. 
%The reqT tool is used in a course based on a specific text book \cite{Lauesen2002} and a specific student project concept \cite{ets170}, and the concepts of the earlier versions of the reqT requirements metamodel reflect that context.

The reqT metamodel includes three types of concepts: entities, attributes and relations. Entities and attributes are nodes in a graph data structure, while relations are edges that can connect entities with sub-graphs. Thus a tree-like structure can be created of arbitrary depth spanning the graph that models some chunk of requirements.

Version 3.0 of the metamodel includes the concepts defined in Appendix A. These concepts and definitions are gatherd from various sources including the IREB Glossary \cite{IREB}, wikipedia, terminology from agile development, variability modelling terminology, and the Lauesen text book \cite{Lauesen2002} used in the RE course at Lund University \cite{ets170}, in which reqT is applied in student role-playing projects.

!!! EXPLAIN EXAMPLE !!!
\begin{spacing}{0.8}
\begingroup
    \fontsize{8pt}{12pt}\selectfont
\begin{lstlisting}
Model(
  Component("apperance") has (
    VariationPoint("color") has (
      Min(0), Max(2),
      Variant("blue"), Variant("red"), Variant("green")),
    VariationPoint("shape") has (
      Min(1), Max(1), Variant("round"), Variant("square")),
    VariationPoint("payment") has (
      Min(1), Max(2), Variant("cash"), Variant("credit")),
    VariationPoint("payment") requires Variant("cash"), 
    Variant("round") excludes Variant("red"),
    Variant("green") requires Variant("square")),
  Component("apperance") requires VariationPoint("shape"), 
  App("free") requires Component("apperance"),  
  App("free") binds (
    VariationPoint("shape") binds Variant("round")),
  App("premium") requires Component("apperance"),  
  App("premium") binds ( 
    VariationPoint("color") binds (Variant("red"), Variant("green")),
    VariationPoint("shape") binds (Variant("round"), Variant("square")),
    VariationPoint("payment") binds Variant("cash")))
\end{lstlisting}
\endgroup
\end{spacing}

\section{Methodology and Data Collection}
!!! explain how sources such as the IREB glossary was used to form the metamodel concept definitions !!!
\section{Data Analysis}

\textbf{Subject background.} The background questions in the survey regards the role of the subject, as shown in Table \ref{table:background}. In summary, the included\footnote{One subject answered NO on all background questions and was therefore excluded.} \input{../analysis/summary.tex}
\begin{table}[H]
\centering
\fontsize{8}{9}\selectfont
\caption{Background of subjects, $N = 15$. The table shows anonymized subject ids}
\label{table:background}
\begin{tabular}{p{0.42\textwidth}| p{0.58\textwidth}}
\textit{Question} & \textit{Subject responding YES}  \\ \hline
\input{../analysis/background.tex}

 \end{tabular}
\end{table}

\textbf{Frequency analysis.} The idea of ''essentiality'' is characterized as the number of subjects that has responded that they (1) use the concepts at least in an informal, non-persistent way, and (2) use the concepts in a similar meaning as in the definition in Appendix A. Table \ref{table:frequency} shows the results of the frequency counts.

%
\begingroup
\setlength{\tabcolsep}{4pt} % Default value: 6pt
\renewcommand{\arraystretch}{1.5} % Default value: 1
\begin{table}[H]
\centering
\fontsize{7}{8}\selectfont
\caption{Frequency analysis of concepts' ''essentiality'', where $n$ is the number of subjects that answered $(use >= 1)$ and $(agree = 2)$ for each concept.}
\label{table:frequency}
\begin{tabular}{l | p{0.33\textwidth} | p{0.30\textwidth} | p{0.25\textwidth}}
\textit{$n$} & \textit{Entities} & \textit{Attributes} & \textit{Relations} \\ \hline
\input{../analysis/essential.tex}
 \end{tabular}
\end{table}
\endgroup
%%%%%%%%%%%%%% DISCUSSION

@@

!!Discuss and conclude!!

{\bf Limitations. } Due to the limited number of subjects and the high degree of homogeneity among subjects with respect to background, it is difficult to analyse and draw conclusions about potential differences in opinions between e.g. teachers and developers. Some subjects needed more time and completed survey offline during the coming days, which may give a variation in how careful the responses were considered. The survey was conducted by the author and inventor of reqT, in conjunction with a seminar and demo of reqT. In order to avoid any positive bias due to advocacy in favour of the reqT metamodel, the survey was held prior to the seminar and demo. This in turn may introduce a threat of limited knowledge among subjects of the idea behind the modelling approach in reqT

{\bf Future work.} %Further directions of research include (1) incorporation of constraints on models for support of prioritization and release planning \cite{Regnell2011}, (2) more elaborate semantic checks to better guide requirements modelers, and (3) graphical visualization of requirements graph models. (4) Natural Language Processing technology including e.g. ambiguity risk detection may be interesting to combine with reqT. (5) It is also important to further investigate the pedagogic advantages and limitations of the approach.

%A major objective of this research preview paper is to expose the latest version of reqT to the community of RE scholars and to invite discussions and contributions.

\subsubsection*{Acknowledgments.} This work is partly funded by  VINNOVA within the EASE project.

@@

%%%%%%%%%%%%%%% APPENDIX WITH DEFINITIONS
\clearpage
%\hspace{-0.9cm}
\section*{Appendix A: Definitions of Metamodel Concepts}
\begin{minipage}{1\linewidth}
\begin{minipage}{0.30\textwidth}

\begin{table}[H]
%\centering
\vspace{-0.8cm}
\fontsize{6}{6.5}\selectfont
%\caption{.}%: Entities, Attributes and Relations.}

\begin{tabular}{|l p{4.5cm}|}
\hline
\textit{Entity} & \textit{Definition}  \\ \hline
\input{../analysis/entity-defs.tex}
\hline
 \end{tabular}
\end{table}
\end{minipage}
 \hspace{4em plus 1fill} 
\begin{minipage}{0.47\textwidth} 
 \begin{table}[H]
%\centering
\fontsize{6}{6.5}\selectfont
%\caption{Definitions of attributes and relations.}
\vspace{-0.8cm}
\label{table-entities}
\begin{tabular}{|l p{4.0cm}|}
 \hline
\textit{Attribute} & \textit{Definition}  \\ \hline
\input{../analysis/attribute-defs.tex}

\hline
\textit{Relation} & \textit{Definition}  \\ \hline
\input{../analysis/relation-defs.tex}
 \hline
\end{tabular}
\end{table}

\end{minipage}
\end{minipage}
